---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis-CI Build Status](https://travis-ci.org/jjchern/zcta.svg?branch=master)](https://travis-ci.org/jjchern/zcta)

# About

This R data package intends to store 2010 Census ZIP Code Tabulation Area (ZCTA) Relationship files. So far in only includes the "2010 ZCTA to County Relationship File" (`zcta_county_rel_10.rda`).

# Motivation

Linking the USPS's ZIP codes to US counties is tedious:
<<<<<<< HEAD

- ZIP codes do not resemble spatial entities; they are created for delivering mails. ZIP codes also change over time.

- To get a truly spatial representations of ZIP codes, the US Census Bureau develops the concept of ZIP Code tabulation areas (ZCTAs), which approximates ZIP codes. But

    - Census does not release an official crosswalk between ZIP codes and ZCTAs.
    
    - Census does release relationship files between ZCTAs and counties, but at least 25% of the ZCTAs cannot be uniquely linked to counties.
=======
    - ZIP codes do not resemble spatial entities; they are created for delivering mails. ZIP codes also change over time.
    - To get a truly spatial representations of ZIP codes, the US Census Bureau develops the concept of ZIP Code tabulation areas (ZCTAs), which approximates ZIP codes. But
        - Census does not release an official crosswalk between ZIP codes and ZCTAs.
        - Census does release relationship files between ZCTAs and counties, but at least 25% of the ZCTAs cannot be uniquely linked to counties.
>>>>>>> 9f8f78c862f157bedba3894e7adf82f565a4a6b0
        
A proposed solution: ZIP codes -> ZCTAs -> counties. This package contains data for connecting the second link. For connecting the first link, see [`zipzcta`](https://github.com/jjchern/zipzcta).

# Useful Links

- [Data Source](https://www.census.gov/geo/maps-data/data/zcta_rel_download.html)
- [Explanation of the 2010 ZCTA to County Relationship File](http://www2.census.gov/geo/pdfs/maps-data/data/rel/explanation_zcta_county_rel_10.pdf)
- [Census ZIP Code Tabulation Areas page](https://www.census.gov/geo/reference/zctas.html)
- [ZIP Code Tabulation Areas (ZCTAs) Delineation Animation](https://www.census.gov/geo/reference/zcta/zcta_delin_anim.html)
- [Overview of the 2010 ZIP Code Tabulation Area (ZCTA) Relationship Files](https://www.census.gov/geo/maps-data/data/zcta_rel_overview.html)
- [2010 ZIP Code Tabulation Area (ZCTA) Relationship File Layouts and Contents](https://www.census.gov/geo/maps-data/data/zcta_rel_layout.html)
- [Census Frequently Asked Questions about ZCTAs](https://ask.census.gov/faq.php?id=5000&rtopic=1795&rsubtopic=8048)

- [The ZIP Code Resources Page](http://mcdc.missouri.edu/allabout/zipcodes.html)
- [All About ZIP Codes: 2010 Supplement](http://mcdc.missouri.edu/allabout/zipcodes_2010supplement.shtml)

# Installation

```r
# install.packages("devtools")
devtools::install_github("jjchern/zcta")
```

# Usage

```{r, message=FALSE, warning=FALSE}
library(dplyr)

# Data
zcta::zcta_county_rel_10

# Show variable labels, and whether value label exists for certain variables
# devtools::install_github("larmarange/labelled")
labelled::var_label(zcta::zcta_county_rel_10)

# Total number of zcta records
nrow(zcta::zcta_county_rel_10)

# Number of distinct zcta
zcta::zcta_county_rel_10 %>% distinct(zcta5) %>% nrow()

# In most instances the ZCTA code is the same as the ZIP Code for an area
# But some zctas fall in more than one county 
# For example, there're 7064 zctas fall in 2 counties
zcta::zcta_county_rel_10 %>% 
  group_by(zcta5) %>% 
  summarise(`Num of counties` = n()) %>% 
  group_by(`Num of counties`) %>% 
  summarise(`Num of zctas` = n())

# To get an one-to-one relationship between zcta and county, assign county to 
# a zcta if the zcta has the most population. For Example:
# Before: zcta 601 fall in county 72001 and 72141
zcta::zcta_county_rel_10 %>% 
  select(zcta5, state, county, geoid, poppt, zpoppct)

# After: relate zcta 601 only to county 72001 as it accounts for 99.43% of the population
one_to_one_pop = zcta::zcta_county_rel_10 %>% 
  select(zcta5, state, county, geoid, poppt, zpoppct) %>% 
  group_by(zcta5) %>% 
  slice(which.max(zpoppct))
one_to_one_pop

# Or assign county to a zcta if the zcta accounts for most of the area.
one_to_one_area = zcta::zcta_county_rel_10 %>% 
  select(zcta5, state, county, geoid, poppt, zpoppct, zareapct) %>% 
  group_by(zcta5) %>% 
  slice(which.max(zareapct))
one_to_one_area

# Merge the two 1 to 1 relationship datasets and identify zctas that have different county match
one_to_one_pop %>% 
  left_join(one_to_one_area, by = "zcta5") %>% 
  select(zcta5, 
         county.x, geoid.x, zpoppct.x, 
         county.y, geoid.y, zpoppct.y) %>% 
  mutate(diff = geoid.x - geoid.y) %>% 
  filter(diff != 0)

# Get county names for the 1 to 1 relationship dataset
# Also keep just states and DC
one_to_one_pop %>% 
  left_join(gaze::county10, by = "geoid") %>% 
  select(zcta5, state, usps, county, geoid, name) %>% 
  filter(state <= 56) %>% 
  group_by(state)
```

